---
title: "Titanic analysis"
output:
  html_document: 
    keep_md: yes
---

Here we are going to fit a classical logistic regression to the Kaggle 
[_Titanic_ data set](https://www.kaggle.com/hesh97/titanicdataset-traincsv).

```{r libs}
lib_vec <- c('readr', 'magrittr', 'ggplot2', 'dplyr', 'jtools')
for (lib in lib_vec){
  library(lib, character.only=TRUE)
}
```

Read in the _Titanic_ data set which I have already downloaded to a local 
location in CSV format.

```{r}
titanic_filename <- "titanic-train.csv"
data_path = "~/repos/titanic/data"
titanic_df <- read_csv(file.path(data_path, titanic_filename), 
                       col_types="iffcfnincncc") %>%
  mutate(age_band = factor(case_when(
    Age < 18 ~ " 1. Under 18",
    Age < 25 ~ " 2. 18 to 24",
    Age < 35 ~ " 3. 25 to 34",
    Age < 45 ~ " 4. 35 to 44",
    Age < 55 ~ " 5. 45 to 54",
    Age < 65 ~ " 6. 55 to 64",
    Age >= 65 ~ " 7. 65 and over",
    TRUE ~ " 8. Other")),
    SibSp = factor(
      case_when(
        SibSp >= 2 ~ "2+",
        TRUE ~ as.character(SibSp)
      )
    ),
    Parch = factor(
      case_when(
        Parch >= 2 ~ "2+",
        TRUE ~ as.character(Parch)
      )
    ), # placeholder
    Cabin = substring(Cabin, 1, 1),
    Cabin = factor(case_when(
      Cabin %in% c('G', 'T') ~ "G/T",
      is.na(Cabin) ~ "Not given",
      TRUE ~ Cabin
    )),
    Fare = factor(case_when(
      Fare <= 10  ~ " 1. (  0,  10]",
      Fare <= 20  ~ " 2. ( 10,  20]",
      Fare <= 30  ~ " 3. ( 20,  30]",
      Fare <= 40  ~ " 4. ( 30,  40]",
      TRUE  ~ " 5. ( 40,    ]",
   )),
   Embarked = factor(
     case_when(
       is.na(Embarked) ~ 'S',
       TRUE ~ Embarked
     )
   )
    )
titanic_df %>% head(20)
```
Now we are going to examine some of this data. The thing we are trying to 
predict--the _target_--is `Survived`; that is, whether the passenger survived
(`1`) or did not (`0`).

```{r}
ggplot(data=titanic_df, aes(x=Survived)) + geom_bar()
```
`SibSp` is the number of siblings or spouse on board.

```{r}
ggplot(data=titanic_df, aes(x=ordered(SibSp))) + geom_bar()
```

`Pclass` is the passenger class.

```{r}
ggplot(data=titanic_df, aes(x=ordered(Pclass))) + geom_bar()
```

`Parch` is the number of parents/children onboard the _Titanic_.
```{r}
ggplot(data=titanic_df, aes(x=ordered(Parch))) + geom_bar()
```
Let's visualise `Fare`.
```{r}
ggplot(data=titanic_df, aes(x=Fare)) + geom_bar()
```

## Fit a GLM

```{r}
glm_01 <- glm(Survived ~ Pclass + Sex + age_band + SibSp + Parch + Fare + Cabin + Embarked,
              data=titanic_df, family=binomial)
anova(glm_01, test='Chisq')
```
```{r}
summary(glm_01)
```
The large effects are on the variables `Sex`, `age_band`, `Cabin` and `Embarked`.
We will also examine `Fare` since there seems to be some signal there.
```{r}
effect_plot(model=glm_01, pred=Sex, interval=TRUE, outcome.scale="link")
effect_plot(model=glm_01, pred=age_band, interval=TRUE, outcome.scale="link")
effect_plot(model=glm_01, pred=Cabin, interval=TRUE, outcome.scale="link")
effect_plot(model=glm_01, pred=Embarked, interval=TRUE, outcome.scale="link")
effect_plot(model=glm_01, pred=Fare, interval=TRUE, outcome.scale="link")


```
We have a broad idea of the factors involved that improved passengers' 
chances of survival on the ill-fated voyage. Let's look now at a couple of 
observations: one each of a strong prediction for survival and a strong 
prediction of the opposite of survival.

```{r}
preds_01 <- predict.glm(glm_01, newdata=titanic_df)
preds_01 %>% max
titanic_max <- preds_01 %>% which.max
titanic_min <- preds_01 %>% which.min

```
First look at the maximum.
```{r}
titanic_predictors <- c('Pclass', 'Sex', 'age_band', 'SibSp', 'Parch',  
    'Fare', 'Cabin', 'Embarked')
titanic_df[titanic_max, titanic_predictors]
glm_01 %>% coef
```

```{r}
0.10831603 + 
  1.39380902 + # Pclass 1
  2.73556786 + # Sex female
  0 + # age_band 1. Under 18
  0 + # SibSp 0
  0.17985123 + # Parch 1
  0.34597961 + # Fare 40+
  0.06157172 + # Cabin B
  0  # Embarked C
preds_max <- preds_01 %>% max
preds_max
1/(1 + exp(-preds_max))
```


```{r}
titanic_predictors <- c('Pclass', 'Sex', 'age_band', 'SibSp', 'Parch',  
    'Fare', 'Cabin', 'Embarked')
titanic_df[titanic_min, titanic_predictors]
glm_01 %>% coef
```


```{r}
0.10831603 + 
  0 + # Pclass 3
  0 + # Sex male
  -3.35951828 + # age_band 7. 65 and over
  0 + # SibSp 0
  0 + # Parch 0
   + # 1. (0, 10]
  -0.84485644 + # Cabin Not given
  -0.47636764  # Embarked S
pred_min <- preds_01 %>% min
pred_min
1/(1 + exp(-pred_min))
```
### Confusion matrix

```{r}
preds <- predict.glm(glm_01, newdata=titanic_df, type='response')
preds %>% head()
thresh <- 0.5
titanic_df$Predict_survived <- ifelse(preds > thresh, 1, 0)
```
```{r}
table(titanic_df$Predict_survived, titanic_df$Survived, dnn=list('Predicted', 'Actual'))
```


## Kaggle submission

As this is a Kaggle data set, we can

1. make predictions against the test set
2. submit the test set to Kaggle to see how well our model performs

Read in the data.
```{r}
test_filename <- "test.csv"
data_path = "~/repos/titanic/data"
test_df <- read_csv(file.path(data_path, test_filename), 
                       col_types="ifcfnincncc") %>%
  mutate(age_band = factor(case_when(
    Age < 18 ~ " 1. Under 18",
    Age < 25 ~ " 2. 18 to 24",
    Age < 35 ~ " 3. 25 to 34",
    Age < 45 ~ " 4. 35 to 44",
    Age < 55 ~ " 5. 45 to 54",
    Age < 65 ~ " 6. 55 to 64",
    Age >= 65 ~ " 7. 65 and over",
    TRUE ~ " 8. Other")),
    SibSp = factor(
      case_when(
        SibSp >= 2 ~ "2+",
        TRUE ~ as.character(SibSp)
      )
    ),
    Parch = factor(
      case_when(
        Parch >= 2 ~ "2+",
        TRUE ~ as.character(Parch)
      )
    ), # placeholder
    Cabin = substring(Cabin, 1, 1),
    Cabin = factor(case_when(
      Cabin %in% c('G', 'T') ~ "G/T",
      is.na(Cabin) ~ "Not given",
      TRUE ~ Cabin
    )),
    Fare = factor(case_when(
      Fare <= 10  ~ " 1. (  0,  10]",
      Fare <= 20  ~ " 2. ( 10,  20]",
      Fare <= 30  ~ " 3. ( 20,  30]",
      Fare <= 40  ~ " 4. ( 30,  40]",
      TRUE  ~ " 5. ( 40,    ]",
   )),
   Embarked = factor(
     case_when(
       is.na(Embarked) ~ 'S',
       TRUE ~ Embarked
     )
   )
    )
test_df %>% head(20)
```
Use the GLM to predict survival.
```{r}
test_preds <- predict.glm(glm_01, newdata=test_df, type='response')
test_preds %>% head()
thresh <- 0.5
test_df$Survived = ifelse(test_preds > thresh, 1, 0)
```

```{r}
outfilename <- "submission_lr.csv"
write.csv(test_df[c("PassengerId", "Survived")], file.path(data_path, outfilename), 
          row.names=FALSE, quote=FALSE)
```

